\Title

### 2012-12-23 synmut.tex
minor edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\DeclareMathOperator\de{d\!}
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\env}{\textit{env}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}

@@

\begin{abstract}
\noindent
Intrapatient HIV evolution is goverened by selection on the protein level in the
arms race with the immune system (killer T-cells and antibodies). Synonymous
mutations do not have an immunity-related phenotype and are often assumed to be
neutral. In this paper, we show that synonymous changes in epitope-rich regions
are often deleterious but still reach frequencies of order one.  We analyze time
series of viral sequences from the V1-C5 part of {\it env} within individual
hosts and observe that synonymous derived alleles rarely fix in the
viral population. Simulations suggest that such synonymous mutations
have a (Malthisuan) selection coefficient of the order of $-0.001$, and that
they are brought up to high frequency by linkage to neighbouring beneficial
nonsynonymous alleles (genetic draft). As far as the biological causes are
concerned, we detect a negative correlation between fixation of an allele and
its involvement in evolutionarily conserved RNA stem-loop structures.
This phenonenon is not observed in other parts of the HIV genome, in which
selective sweeps are less dense and the genetic architecture less constrained.
\end{abstract}

@@

\section{Introduction}
HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(AB)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
AB response against a particular HIV epitope, mutations in the viral genome that reduce or prevent
recognition of the epitope frequently emerge. Escape mutations in epitopes
targeted by CTLs typically evolve during early infection and spread rapidly through
the population~\citep{mcmichael_immune_2009}. During chronic infection, the most
rapidly evolving part of the HIV genome are the so called variable loops of the
envelope protein gp120, which need to avoid recognition by neutralizing ABs.
Mutations in \env~, the gene encoding for gp120, spread through the population
within a few months (see \figurename~\ref{fig:aft}, solid lines).
The (Malthusian) effect size of these beneficial mutations is of the
order of $s_a \sim 0.01$~\citep{neher_recombination_2010}.

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximate neutral markers in studies of viral evolution. Neutral
markers are very useful in practice, because they can be used to make inferences
about the stochastic forces driving evolution~\citep{yang_statistical_2000}.
The viral genome, however, needs to satisfy further constraints in addition to
immune escape, such as efficient processing and translation, nuclear export, and
packaging into the viral capsid: all these processes operate at the RNA level
and are sensitive to synonymous changes. A
few functionally important RNA elements are well characterized. For example, a certain RNA
sequence, called \rev{} response element (RRE), is used by HIV to enhance
nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
well studied case is the interaction between viral reverse transcriptase, viral
ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming
reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure
in the viral 5' untranslated region~\citep{barat_interaction_1991,
paillart_vitro_2002}. Furthermore, recent studies have shown that genetically
engineered HIV strains with skewed codon usage bias (CUB) patterns towards more
or less abundant tRNAs replicate better or worse,
respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}.
Purifying selection beyond the protein sequence is therefore expected, while it
seems reasonable that the bulk of positive selection through the immune system
should be restricted to amino acid sequences.

INFLUENZA PSEUDO VACCINE.

%SYNONYMOUS CONSERVATION. DO WE HAVE A PLOT OF GENOME WIDE CONSERVATION, MAYBE
%FOR SUPPLEMENT? YES

Here, we characterize the dynamics of synonymous mutations in \env{} and show
that a substantial fraction of these mutations are deleterious. We further show
that, although such synonymous mutations cannot be used as neutral markers, the
degree to which they hitchhike with nearby nonsynonymous mutations is very
informative. Their ability to hitchhike for extended times, which is a core
requirement for our analysis, is rooted in the small recombination rate of
HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
analysis of fixation probabilities to the nonsynonymous mutations, we show that
time dependent selection or strong competition of escape mutations inside the
same epitope are necessary to explain the observed patterns of fixation and
loss.

@@

\FIG{aft} shows time series data of the frequencies of all mutations observed
\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999,
liu_selection_2006}. Despite many synonymous mutations reaching high frequency,
very few fix. This observation is quantified in panels \ref{fig:fixp1} and
\ref{fig:fixp2}, which stratify the data of 7-10 patients according to the
frequency at which different mutations are observed (see methods). Considering
all mutations in a frequency interval around $\nu_0$ at some time $t_i$, we
calculate the fraction that is found at frequency 1, at frequency 0, or at
intermediate frequency at later times $t_f$. Plotting these fixed, lost, and
polymorphic fraction against the time interval $t_f-t_i$, we see that most
synonymous mutations segregate for roughly one year and are lost much more
frequently than expected. The long-time probability of fixation versus
extinction is shown as a function of the initial frequency $\nu_0$ in
panel~\ref{fig:fixp2}. In contrast to synonymous mutations, the nonsynonymous
seem to follow more a less the neutral expectation -- a point to which we will
come back below. 
\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8}
\caption{Synonymous mutations rarely fix in \env, C2-V5: mutation frequency
 trajectories observed in patient 8~\cite{shankarappa_consistent_1999};
 Nonsynonymous and synonymous mutations are shown as solid and dashed lines,
 respectively. Colors indicate the position of the site along the C2-V5 region
 (red to blue) MAYBE MAKE FIGURE WITH SYNONYMOUS AND NONSYN
 SEPARATELY. While nonsynonymous mutations frequently fix, very few synonymous
mutations do even though they are frequently observed at intermediate
frequencies.}
\label{fig:aft}
\end{center}
\end{figure}

@@

These observations suggest that many of the synonymous polymorphisms in the part
of \env~that includes the hypervariable regions are deleterious, while outside
this regions polymorphisms are mostly roughly neutral.

@@

\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are  secondary structures in the viral RNA. If any RNA
secondary structures are relevant for HIV replication, mutations in nucleotides
involved in those base pairs are expected to be deleterious and to revert
preferentially. Many functionally important secondary structure elements have
been characterized, including  the RRE~\citep{fernandes_hiv-1_2012} the
5' UTR pseudoknot interacting with thehost tRNA$^\text{Lys3}$~\citep{barat_interaction_1991,
paillart_vitro_2002}. It has been suggested early on that parts of the viral
genome that has the potential to form stems is better conserved that the
remainder~\citep{forsdyke_reciprocal_1995}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\env}{\textit{env}}

\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\DeclareMathOperator\de{d\!}
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\shankaregion{C2-C5}

@@

\begin{abstract}
\noindent
Intrapatient HIV evolution is governed by selection on the protein level in the
arms race with the immune system (killer T-cells and antibodies). Synonymous
mutations do not have an immunity-related phenotype and are often assumed to be
neutral. In this paper, we show that synonymous changes in epitope-rich regions
are often deleterious but still reach high frequencies.  We analyze time
series of viral sequences from the \shankaregion~part of {\it env} within individual
hosts and observe that synonymous derived alleles rarely fix in the
viral population. Simulations suggest that such synonymous mutations
have a (Malthusian) selection coefficient of the order of $-0.001$, and that
they are brought up to high frequency by hitchhiking with neighboring beneficial
nonsynonymous alleles. We detect a negative correlation between fixation of an allele and
its involvement in RNA stem-loop structures.
Deleterious synonymous mutations are observed in other parts of the HIV genome, in which
selective sweeps are less dense and hitch-hiking not as strong.
\end{abstract}

@@

\section{Introduction}
HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(AB)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
AB response against a particular HIV epitope, mutations in the viral genome that reduce or prevent
recognition of the epitope frequently emerge. Escape mutations in epitopes
targeted by CTLs typically evolve during early infection and spread rapidly through
the population~\citep{mcmichael_immune_2009}. During chronic infection, the most
rapidly evolving part of the HIV genome are the so called variable loops of the
envelope protein gp120, which need to avoid recognition by neutralizing ABs.
Mutations in \env, the gene encoding for gp120, spread through the population
within a few months (see \figurename~\ref{fig:aft}, solid lines).
The (Malthusian) effect size of these beneficial mutations is of the
order of $s_a \sim 0.01$~\citep{neher_recombination_2010}.

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximate neutral markers in studies of viral evolution. Neutral
markers are very useful in practice, because they can be used to make inferences
about the stochastic forces driving evolution~\citep{yang_statistical_2000}.
The viral genome, however, needs to satisfy further constraints in addition to
immune escape, such as efficient processing and translation, nuclear export, and
packaging into the viral capsid: all these processes operate at the RNA level
and are sensitive to synonymous changes. A
few functionally important RNA elements are well characterized and show substantially 
less synonymous variation across the HIV population (see Fig.~S1). For example, a certain RNA
sequence, called \rev{} response element (RRE), is used by HIV to enhance
nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
well studied case is the interaction between viral reverse transcriptase, viral
ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming
reverse transcription (RT) and bound by a pseudoknotted RNA structure
in the viral 5' untranslated region~\citep{barat_interaction_1991,
paillart_vitro_2002}. Furthermore, recent studies have shown that genetically
engineered HIV strains with skewed codon usage bias (CUB) patterns towards more
or less abundant tRNAs replicate better or worse,
respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}.
Similarly, influenza can be dramatically attenuated by reducing its translation efficiency \citep{pseudo_vaccine}.
Purifying selection beyond the protein sequence is therefore expected, while it
seems reasonable that the bulk of positive selection through the immune system
should be restricted to amino acid sequences.

Here, we characterize the dynamics of synonymous mutations in \env{} and show
that a substantial fraction of these mutations are deleterious. We further show
that, although such synonymous mutations cannot be used as neutral markers, the
degree to which they hitchhike with nearby nonsynonymous mutations is very
informative. Their ability to hitchhike for extended times is rooted in the small recombination rate of
HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
analysis of fixation probabilities to the nonsynonymous mutations, we show that
time dependent selection or strong competition of escape mutations inside the
same epitope are necessary to explain the observed patterns of fixation and
loss.

@@

\FIG{aft} shows time series data of the frequencies of all mutations observed
\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999,
liu_selection_2006}. Despite many synonymous mutations reaching high frequency,
very few fix. This observation is quantified in panels \ref{fig:fixp1} and
\ref{fig:fixp2}, which stratify the data of 7-10 patients according to the
frequency at which different mutations are observed (see methods). Considering
all mutations in a frequency interval around $\nu_0$ at some time $t_i$, we
calculate the fraction that is found at frequency 1, at frequency 0, or at
intermediate frequency at later times $t_f$. Plotting these fixed, lost, and
polymorphic fraction against the time interval $t_f-t_i$, we see that most
synonymous mutations segregate for roughly one year and are lost much more
frequently than expected if they were neutral. The long-time probability of fixation versus
extinction is shown as a function of the initial frequency $\nu_0$ in
panel~\ref{fig:fixp2}. In contrast to synonymous mutations, the nonsynonymous
seem to follow more a less the neutral expectation -- a point to which we will
come back below. 
\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8}
\caption{Synonymous mutations rarely fix in \env, C2-V5: mutation frequency
 trajectories observed in patient 8~\cite{shankarappa_consistent_1999};
 Nonsynonymous and synonymous mutations are shown as solid and dashed lines,
 respectively. Colors indicate the position of the site along the C2-V5 region
 (red to blue) MAYBE MAKE FIGURE WITH SYNONYMOUS AND NONSYN
 SEPARATELY. While nonsynonymous mutations frequently fix, very few synonymous
mutations do even though they are frequently observed at intermediate
frequencies.}
\label{fig:aft}
\end{center}
\end{figure}

@@

These observations suggest that many of the synonymous polymorphisms in the part
of \env~that includes the hypervariable regions are deleterious, while outside
this regions polymorphisms are mostly roughly neutral. Note that this does not imply that all 
synonymous mutations are neutral -- only those mutations observed at high frequencies tend 
to be neutral.

@@

\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are  secondary structures in the viral RNA. If any RNA
secondary structures are relevant for HIV replication, mutations in nucleotides
involved in those base pairs are expected to be deleterious and to revert
preferentially. Indeed, synonymous variation across the HIV population is variable, 
suggesting purifying selection beyond the protein coding level.
